---
title: "timeseries-example"
output: html_document
vignette: >
  %\VignetteIndexEntry{timeseries-example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tidyverse)
devtools::load_all()
# library(growthrates)
here::i_am("vignettes/timeseries-example.Rmd")
source(here::here("vignettes/vignette-utils.R"))
ggplot2::theme_set(ggplot2::theme_minimal())
```

# Line lists and time series

ID data either comes as a set of individual observations with some sort of 
timestamp (a line list) or as a count of events happening within a specific 
period (day, week, month etc.) as an incidence tmie series.

For both these data types there may also be a class associated with each 
observation which could be for example the variant of a virus, or a particular
test type, for example. It may make sense to compare the relative growth of the
different classes.

```{r}
dates = seq(as.Date("2020-01-01"),by=7,length.out = 52)

as.time_period(dates)
```
