---
title: "England COVID-19 cases"
output: html_document
vignette: >
  %\VignetteIndexEntry{England COVID-19 cases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
here::i_am("vignettes/covid-timeseries.Rmd")
source(here::here("vignettes/vignette-utils.R"))
library(tidyverse)
library(growthrates)
```

# Incidence and growth rate from case positive counts

This method uses a case positive count dataset bundled with growth rates which
is age stratified (age grouping is in the `class` column). We will look at the
age stratification in different vignettes but in this instance we want to
aggregate it into an England wide rate. This is the purpose of `time_aggregate()`
which performs a simple summarisation.

```{r}
england_covid %>% glimpse()

tmp = growthrates::england_covid %>%
  time_aggregate(count=sum(count))
```

The raw covid case count on a `log1p` scale is the total detected cases per day.


```{r}
fit = tmp %>%
   poisson_locfit_model() 

plot_incidence(fit,raw = tmp, colour="blue",size=0.025)+
  scale_y_log1p(n=7)
```

Major events in the timeseries can be plotted on the axes. We've focussed on the first 2 
years of the pandemic:

```{r}
plot_incidence(fit, raw = tmp,events = england_events, colour="blue",size=0.025)+
  scale_y_log1p(n=7) + coord_cartesian(xlim=as.Date(c("2020-01-01","2022-01-01")))

```

The incidence model assumes case rates result from a Poisson process and the
rate is estimated with a time varying locally fitted polynomial with degree
defined by the `deg` parameter, using a log link function, according to the
methods of Loader et al. (see `citation("locfit")`). The fitting process is a
local maximum likelihood estimation and uses a bandwidth defined to account for
the data points within `window` of the time point being estimated.

The gradient of the fitted polynomial on the log scale, is the exponential
growth rate. This is a scale independent view of the rate of growth of the
epidemic. This estimation methodology is compared to consensus estimates from
the SPI-M-O UK government advisory group in red, shifted forward in time by 21
days. The SPI-M-O estimates made during the pandemic were retrospective whereas
these ones can use information from before and after the time point and now may
better represent the timing of changes.

```{r}

plot_growth_rate(fit,events = england_events, colour="blue")+
  coord_cartesian(xlim=as.Date(c("2020-01-01","2022-01-01")), ylim=c(-0.15,0.15))+
  geom_errorbar(data=england_consensus_growth_rate,aes(x=date-21,ymin=low,ymax=high),colour="red")

```

The state of the epidemic is described by both incidence and growth, and the phase plots
allow us to see both at different time points. In this case the epidemic state in 
the 10 weeks leading up to Christmas in 2021, 2022 and 2023:

```{r}


plot_growth_phase(fit,
    timepoints = as.Date(c("Xmas 2020"="2020-12-25","Xmas 2021"="2021-12-25","Xmas 2022"="2022-12-25")),
    duration = 70, 
    interval = 7,
    colour="blue"
)

```

# Reproduction estimation from growth rates

The growth rate has a unit or "per day" in this example. From it we can derive
the reproduction number. Using the methods of Wallinga and Lipsitch and an
estimate of the infectivity profile of COVID-19. This describes the probability
that an infectee is infected x days after the infector, and includes a temporal
dimension rendering the reproduction number a dimensionless quantity reflecting 
the average number of infectees resulting from each infector.

`growthrates` has an estimate of the infectivity profile based on a
meta-analysis of serial interval estimates of COVID-19. The infectivity profile
is a bootstrapped set of discrete probability distributions. It is truncated at
14 days.

```{r}



ggplot2::ggplot()+
  geom_violin(
    data = growthrates::covid_infectivity_profile %>% tidyr::complete(time=0:max(time), fill = list(probability=0)),
    mapping = ggplot2::aes(x=as.factor(time),y=probability),
    colour="blue",
    alpha=0.1
  )+
  geom_line(
    data = growthrates::covid_infectivity_profile %>% 
      tidyr::complete(time=0:max(time), fill = list(probability=0)) %>%
      dplyr::group_by(time) %>%
      dplyr::summarise(m = mean(probability)) %>% 
      dplyr::ungroup(),
    mapping = ggplot2::aes(x=as.factor(time),y=m, group=1),
    inherit.aes = FALSE
  )


```

For each growth rate estimate this methods uses 1000 bootstraps to propagate
uncertainty and is hence somewhat slow. Here we use `memoise` to cache the
result. As with before effective $R_t$ estimates are compared to consensus
values from the SPI-M-O group (red):

```{r}

.cache = memoise::cache_filesystem(rappdirs::user_cache_dir("growthrates"))

cached_rt_from_growth_rate = memoise::memoise(
  growthrates::rt_from_growth_rate,
  cache = .cache
)


rt_fit = fit %>% cached_rt_from_growth_rate(ip = covid_infectivity_profile)

plot_rt(rt_fit, events = england_events, colour="blue")+
  coord_cartesian(xlim=as.Date(c("2020-01-01","2022-01-01")), ylim=c(0.6,1.6))+
  geom_errorbar(data=england_consensus_rt,aes(x=date-21,ymin=low,ymax=high),colour="red")
```

# Prevalence and growth rate from test positivity rates

Test availability was not consistent during the pandemic. In the early stages 
PCR tests were difficult to obtain and case positive incidence estimates are
thought to be a vast underestimate. During cetain parts of the pandemic targetted
testing of high risk groups occurred. Test positivity is a different view
on the pandemic and accounts for some of these biases and introduces others 
of its own. For this the data must contain a `denom` column which in this
case represents the number of tests conducted:

```{r}
england_covid_pcr_positivity %>% glimpse()


```


```{r}

fit2 = england_covid_pcr_positivity %>% 
  growthrates::proportion_locfit_model()

plot_proportion(fit2, england_covid_pcr_positivity, events = england_events, size=0.25, colour="blue")+
  coord_cartesian(xlim=as.Date(c("2020-01-01","2022-01-01")))

```

In this case the gradient of the proportion on the logistic scale is an estimate
of the growth rate. It is in some senses relative to the growth of the testing
effort but in this case produces an answer very similar to the incidence model.

```{r}

plot_growth_rate(fit2, events = england_events, colour="blue")+
  coord_cartesian(xlim=as.Date(c("2020-01-01","2022-01-01")), ylim=c(-0.15,0.15))+
  geom_errorbar(data=england_consensus_growth_rate,aes(x=date-21,ymin=low,ymax=high),colour="red")
```

With similar growth rate estimates this method can also theoretically be used to
calculate estimates of $R_t$. Growth-proportion phase diagrams can also compare
different points in times or as we will see elsewhere, between different
populations.

```{r}

plot_growth_phase(fit2,
    timepoints = as.Date(c("Xmas 2020"="2020-12-25","Xmas 2021"="2021-12-25","Xmas 2022"="2022-12-25")),
    duration = 70, 
    interval = 7,
    colour="blue"
)

```
