---
title: "Population comparisons and incidence"
output: html_document
vignette: >
  %\VignetteIndexEntry{Population comparisons and incidence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tidyverse)
library(patchwork)
# devtools::load_all()
library(growthrates)
here::i_am("vignettes/incidence-trends.Rmd")
source(here::here("vignettes/vignette-utils.R"))
```

Comparisons between epidemic states in different populations separated by for example
geography, or in this case age, requires a normalisation by the population size
in either absolute or proportional terms. Based on the 2021 census we have some
demo data for the demographics of England:

```{r}
glimpse(england_demographics)
```

# Incidence Poisson rate model

A plot of the normalised incidence rates of COVID-19 by population size, shows 
initially the rate of COVID cases was highest in the elderly. By late 2020 this
pattern had changed and rates were uniform accross age groups. In early 2021
as vaccination took hold and school testing was rolled out, younger age groups
had higher rates of COVID positive tests, with a curious spike in the very young
age groups around November 2021. In early 2022 the pattern reversed and the 
elderly again became the age group with the highest rates, and that pattern has
persisted until the present.

```{r}

tmp = growthrates::england_covid %>%
   growthrates::poisson_locfit_model(window=21) %>%
   growthrates::normalise_incidence(growthrates::england_demographics)

raw_pop = growthrates::england_covid %>% inner_join(england_demographics, by="class")

plot_incidence(tmp,raw = raw_pop, size=0.25)+scale_y_log1p(n=7)+
  scale_colour_viridis_d(aesthetics = c("fill","colour"))
  
```

The use of test positives as a proxy for COVID incidence is clearlly potentially
biased by testing (partilcularly in the first wave where testing was limited to those
in hospital). A more reliable comparison in this situation would be a test
positivie proportion, but unfortunately testing rates are not published broken 
down by age.

The exponential growth rate is already normalised by population size. Comparisons
of the growth rate in these populations gives an idea of how tightly coupled they are.
In most age groups the epidemic is growing and shrinking in sync apart from possibly
the very young. COVID detections in this age group were not particularly reliable though
and this would be easy to over-interpret.

```{r}


plot_growth_rate(tmp)+
  scale_fill_viridis_d(aesthetics = c("fill","colour"))+
  coord_cartesian(ylim=c(-0.15,0.15))



```

The combination of growth and normalised incidence allows us to compare the
epidemic state at different time points, in this case Christmas day in 2020,
2021 and 2022. This shows the same data as the previous graphs.

```{r}

plot_growth_phase(tmp,
    timepoints = as.Date(c("Xmas 2020"="2020-12-25","Xmas 2021"="2021-12-25","Xmas 2022"="2022-12-25")),
    duration = 70, 
    interval = 7
)+
  scale_colour_viridis_d()

```


# Proportion model

There are two possible proportions models which woudl be of interest here. As
mentioned about the proportion of positive tests in each age group would give
us a clearer picture of whether the differences between age groups are down to 
differential testing, but unfortunately not available in this data set.

The second potential use is the distribution of ages in the test positive age
group. This age distribution gives us some information about where the burden of 
disease is in the population but is also biased by test prioritisation.

A multinomial proportion shows similar patterns as the normalised incidence plot
above:

```{r}


tmp2 = growthrates::england_covid %>%
   growthrates::proportion_locfit_model(window=21)

p1 = plot_multinomial(tmp2,normalise = TRUE)+
  scale_fill_viridis_d()

p2 = ggplot(england_demographics)+
  geom_bar(aes(x="baseline",y=population/sum(population)*100,fill=class), stat="identity", position="stack", colour="black", linewidth=0.1)+
  scale_fill_viridis_d(guide="none")+
  xlab(NULL)+
  ylab(NULL)+
  theme(axis.text.y = element_blank())+
  ggplot2::coord_cartesian(expand=FALSE)

p1+p2+patchwork::plot_layout(nrow=1,widths = c(20,1),guides = "collect")

```

The age distribution of test positives can be normalised by the age distribution
of the population. This give us a relative proportion of age groups in people
testing positive versus expected in the population. This is conceptually a
relative risk but of age group given COVID status 

i.e. it is $\frac{P(age = 80+|COVID+)}{P(age = 80+)}$

At any point in time for a given population this quantity is centred around 1
and comparing it to growth rate gives us a possibly clearer picture of the
trajectory of the relative distribution of COVID in the population. In Xmas 2021
although the majority of cases was in the young, relatively high growth in the
elderly population meant that they were catching up, and from above we can see
that by early 2022 the elderly had highest COVID positive rates. In 2022
however, the separation of the age groups was established and the trajectories
were acting to preserve that separation.

```{r}

tmp3 = tmp2 %>% normalise_proportion(england_demographics)

plot_growth_phase(tmp3,
    timepoints = as.Date(c("Xmas 2020"="2020-12-25","Xmas 2021"="2021-12-25","Xmas 2022"="2022-12-25")),
    duration = 70, 
    interval = 7
)+
  scale_colour_viridis_d()

```



